---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# sqlr

<!-- badges: start -->
[![R-CMD-check](https://github.com/nbenn/sqlr/actions/workflows/check-standard.yaml/badge.svg)](https://github.com/nbenn/sqlr/actions/workflows/check-standard.yaml)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

Describe a database schema in R, emit it as data definition language for a
given SQL dialect, and read it back out of a live database.

Generating SQL from R is a solved problem in patches. DBI carries type
mappings and quoting, dbplyr covers the query side, and packages that need a
`CREATE TABLE` tend to assemble one by hand. What is missing is the middle
term: a representation of a schema that is neither a string nor a connection,
which you can build, render, reflect and compare. That is what this package
is for.

## Installation

``` r
pak::pkg_install("nbenn/sqlr")
```

The core package renders nothing on its own -- a dialect supplies the
spellings. Install whichever you need:

``` r
pak::pkg_install("nbenn/sqlr.postgres")
pak::pkg_install("nbenn/sqlr.sqlite")
```

## Describing a schema

Tables, columns, constraints and indexes are ordinary constructor calls, and
nest in the order you would write them.

```{r schema}
library(sqlr)
library(sqlr.sqlite)

schema <- sqlr_schema(
  "main",
  sqlr_table(
    "users",
    sqlr_column("id", sqlr_bigint(), null = FALSE),
    sqlr_column("email", "varchar(255)", null = FALSE),
    sqlr_column("nickname", sqlr_text(), default = "anon"),
    sqlr_primary_key("id"),
    sqlr_unique("email")
  ),
  sqlr_table(
    "orders",
    sqlr_column("id", sqlr_bigint(), null = FALSE),
    sqlr_column("user_id", sqlr_bigint(), null = FALSE),
    sqlr_column("total", sqlr_numeric(10, 2)),
    sqlr_primary_key("id"),
    sqlr_foreign_key("user_id", "users", "id", on_delete = "cascade"),
    sqlr_index("user_id", name = "orders_user_idx")
  )
)
```

Types are given either as objects or as their SQL-standard spelling, so
`sqlr_varchar(255)` and `"varchar(255)"` mean the same thing. Check
constraints go in the same way, via `sqlr_check("total > 0")`.

## Rendering

```{r render}
cat(paste0(sqlr_render(schema, sqlite()), ";"), sep = "\n\n")
```

Rendering a schema is not rendering each table in turn. Foreign keys can form
cycles -- a self-reference, or two tables pointing at one another -- so tables
come out in dependency order, and a key that would close a cycle is lifted
into a trailing `ALTER TABLE`. Engines without `ADD CONSTRAINT`, SQLite among
them, opt out of that and keep every constraint inline.

## Reading a schema back

Point `sqlr_reflect()` at a connection and it returns the same representation
`sqlr_render()` consumes, built from the database catalogue rather than by
parsing SQL.

```{r reflect}
con <- DBI::dbConnect(RSQLite::SQLite(), ":memory:")

for (stmt in sqlr_render(schema, sqlite())) DBI::dbExecute(con, stmt)

sqlr_equal(schema, sqlr_reflect(con))
```

That round trip is the package's own test oracle, and it is stricter than it
looks. Comparison is structural: constraints and indexes are matched on what
they constrain rather than on their order or their names. A schema you wrote
therefore compares equal to one reflected out of a database that picked its
own constraint names, which is what makes comparing against an existing
database useful rather than a wall of spurious differences.

How much survives the trip is bounded by what a database will tell you.
Postgres reports check constraints, so they compare; SQLite records them only
inside the original `CREATE TABLE` text, so reflecting them would mean parsing
DDL and the dialect declines to guess. Each dialect documents where it stops.

When two schemas do differ, `sqlr_diff()` says how.

```{r diff}
altered <- sqlr_table(
  "users",
  sqlr_column("id", sqlr_int(), null = FALSE),
  sqlr_column("email", "varchar(255)"),
  sqlr_primary_key("id")
)

sqlr_diff(schema@tables[[2L]], altered)
```

```{r cleanup, include = FALSE}
DBI::dbDisconnect(con)
```

## Dialects

A dialect is an object, not a connection. Rendering therefore needs no
database and no driver package installed, and the test suite runs without a
server. Given a live connection, `sqlr_for()` resolves the right dialect.

There is deliberately no generic fallback dialect. Nothing renders "ANSI SQL",
because no engine implements it faithfully and an untestable default would
quietly accumulate behaviour no database had ever accepted. A connection with
no dialect registered is an error rather than a guess.

Writing one means implementing the protocol in `?sqlr_render` and
`?sqlr_reflect`: how types spell in both directions, and whatever the engine
does differently.

## Scope

Currently modelled: schemas, tables, columns, types, primary keys, unique and
foreign key constraints, check constraints, indexes, defaults and identity
columns.

Not yet: views, triggers, sequences beyond identity, partitions and grants.
Types a dialect does not model survive as `sqlr_other()` and render verbatim,
so an unsupported column does not stop you reflecting the table around it.

Comparing two schemas is a step towards generating the migration between
them. The representation is built with that in mind -- object identity and
equality are defined so a diff is possible -- but the change engine itself is
not written yet.
